---
title: "REvenT workflow (ENG)"
author: "Federico M. Stefanini"
date: " August 2020 - rel 1.5.7"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REvenT workflow (ENG)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


# Introduction

This document elaborates an example on how to work with
package *REvenT*, a suite of functions to make probability calculus using
trees of events.

**NOTE**: this package has not been released on CRAN.
It might contain functions that will not be released in the final version of this package.
Extensive testing is in progress.


# A Simple case study

Below,  three partitions of a sample space are considered:

```{r}
library(REvenT)
testPart2 <- list(Pa1= c("sane","sick","dead"),
                  Pa2= c("cicatrix","normal"),
                  Pa3=c("young","adult","old"))
testPart2
```
where *Pa* indicates what  partition of events is introduced,
say from *Pa1* on.

The first step is the creation of a dataframe, in which all paths from root
to leaves in the given tree are listed :
```{r}
primaColPat <- REvenT::eve_tree(testPart2)
primaColPat
```

Note the attribute "Tipologia" of this object:
```{r}
attr(primaColPat,"Tipologia")
```


We can now create a matrix that contains conditional probability values:
```{r}
primoPTree<- proba_tree(primaColPat)
primoPTree
```
where each row-path made by events corresponds
to a row in the matrix of probability values.


Conditional probability values have to be provided. 
Two main ways are defined to enter probability values by
using  helper functions.
In the first way,   a string is defined 
for representing the event, besides providing  a probability value:   
"Pa\<i\>:\<event\>;"    
where all terms between angular brackets 
must be substituted like shown in the example below.
In "Pa\<i\>" an integer number indicating which partition is
considered must be inserted, thus we have:
```{r}
test1 <-'Pa1:sick;Pa2:normal'
tmp7 <-assignPro(primoPTree,test1,0.23)
tmp7$proba_tree
```
where it clearly happens that all paths interested by this event are
simultaneously assigned.



A second  way to perform this step  is to use a more general helper function.
It  firstly enumerates all the required assignements and thus
repetitions can be avoided:
```{r}
tmpGen <- genProbLs(primoPTree)
tmpGen[,-1]
```

It is convenient to indicate all paths and events making 
esplicit the partition they belong to.
For example, let's consider the following values:
```{r}
cbind(tmpGen$event[1:3])
tmpGen$probVal[1:3] <- c(0.1, 0.7, 0.2)
tmpGen[1:3,]
```


It is mandatory to check that coherence of probability assignements 
is present, e.g. the sample space must have probability equal to $1$:
```{r}
sum(tmpGen[1:3,4])
```

Further assignementsin this example, after elicitation, might be:
```{r}
tmpGen$probVal[c(4,7)] <- c(0.1, 0.9)
tmpGen[c(4,7),-1]

tmpGen$probVal[c(5,8)] <- c(0.3, 0.7)
tmpGen[c(5,8),-1]

tmpGen$probVal[c(6,9)] <- c(0.95, 0.05)
tmpGen[c(6,9),-1]

tmpGen$probVal[c(10,16,22)] <- c(1,5,10)/16
tmpGen[c(10,16,22),-1]

tmpGen$probVal[c(10,16,22)+1] <- c(5,5,5)/15
tmpGen[c(10,16,22)+1,-1]

tmpGen$probVal[c(10,16,22)+2] <- c(25,5,15)/45
tmpGen[c(10,16,22)+2,-1]

tmpGen$probVal[c(10,16,22)+3] <- c(50,5,25)/80
tmpGen[c(10,16,22)+3,-1]

tmpGen$probVal[c(10,16,22)+4] <- c(25,5,25)/55
tmpGen[c(10,16,22)+4,-1]

tmpGen$probVal[c(10,16,22)+5] <- c(1,5,10)/16
tmpGen[c(10,16,22)+5,-1]
```

The overall set of values is shown below:
```{r}
tmpGen[,-1]
```



A helper function is available to fetch   such values
into the original *eve\_tree* object:
```{r}
primoPTree <- fetchProb(primoPTree,tmpGen)
primoPTree$proba_tree
```





# Marginalization and conditioning


Firstly, probability values of the conjunction of events along all root-to-leave
paths are calculated:
```{r}
tavolaCongiunta <- congiungi(primoPTree)
tavolaCongiunta
```
thus we can check the value implied 
for the sure event:
```{r}
sum(tavolaCongiunta$jointPro)
```
as expected.
 

Now the marginalization is possible
with respect to some events in their own partition, for example:
```{r}
mymargin1 <- marginalizza(tavolaCongiunta,"Pa2:cicatrix;Pa3:adult")
mymargin1
```

```{r}
marginalizza(tavolaCongiunta,"Pa2:cicatrix")
marginalizza(tavolaCongiunta,"Pa1:dead;Pa3:old")
```


Last, it is possible to calculate   conditional probability
values, i.e. with Bayes' rule.
<!--  
```{r}
estraSottotavola(tavolaCongiunta,"Pa2:cicatrix;Pa3:adult")
```
thus after conditioning:
```{r}
myCondizio <-estraSottotavola(tavolaCongiunta,
                              "Pa2:cicatrix;Pa3:adult")
myCondizio$jointPro <- myCondizio$jointPro / mymargin1$probaMargin
cbind(
      myCondizio[,1:4],
      condizionata= myCondizio$jointPro/ mymargin1$probaMargin
)
```


-->

It is convenient to use a function that performs the above calculations
in one step (invocation), starting from a probabilistic definition
based on strings.
Note that the estractor does not depend on the order of conditioning events:
```{r}
estraSottotavola(tavolaCongiunta,"Pa3:adult;Pa1:sane;Pa2:normal")
estraSottotavola(tavolaCongiunta,"Pa2:normal;Pa3:adult;Pa1:sane")
```


A function  called *condiziona* makes all intermediate calculations and it only shows  results:
```{r}
tetCondiz <- condiziona(tavolaCongiunta,
   "Pa1:sane  | Pa3:old; Pa2:cicatrix")
tetCondiz
```

It could be the case that all the distribution is needed, thus in such a case:
```{r}
condiziona(tavolaCongiunta, "   | Pa3:old; Pa2:cicatrix")
```
that adds to one.



Even simpler conditioning does work, and we can ckeck that they add to one:
```{r}
condiziona(tavolaCongiunta, "   | Pa2:cicatrix")
sum(condiziona(tavolaCongiunta, "   | Pa2:cicatrix")$condPro)
```


The marginal distribution after conditioning is:
```{r}
condiziona(tavolaCongiunta, "    | Pa3:adult")
condiziona(tavolaCongiunta, " Pa2:normal   | Pa3:adult")
```

In case the events of the third partition are not observed, we have:
```{r}
test7 <-condiziona(tavolaCongiunta, " Pa1:sane   | Pa2:normal")
test7
```
but here a further marginalization is needed:
```{r}
marginalizza(test7,"Pa1:sane;Pa2:normal")$tavolaMarginale
```





